---
title: "Likelihood figures"
author: "SrenBlikdal"
date: "2026-06-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(bsseq)
library(tidyverse)
library(rtracklayer)
library(GenomicRanges)
library(patchwork)
library(gghalves)
```

# Benchmarking likelihood framework on HG002 data

## Obtain publicly available HG002 ONT data  

### Download data from AWS S3 bucket:

```{r cars}
summary(cars)
```

## Generate HG002 truthset for benchmarking

Download HG002 vcf and bed files from GIAB truth set:
s3://ont-open-data/giab_2023.05/analysis/benchmarking/truthset/

https://42basepairs.com/browse/s3/ont-open-data/giab_2023.05/analysis/benchmarking/truthset?file=HG002_GRCh38_1_22_v4.2.1_benchmark.vcf.gz

Import vcf file with with HG002 variants for GRch38 and bed files high confidence regions i.e. no inconsistent calls between different technologies.
```{r}
#import the vcf file
vcf<-read_tsv(
    "~/Documents/CpG_likelihoods/HG002_GRCh38_1_22_v4.2.1_benchmark_chr1.vcf", 
    comment = "#", 
    col_names = FALSE
)

colnames(vcf)<-c(
    "ch","start","X","ref","alt","Y","Z","info", "tag", "tag_info"
) 
vcf$ch<-gsub("chr1","NC_000001.11",vcf$ch)

# read high confidence regions and format them as GRanges
chr1_gr <- import(
    "~/Documents/CpG_likelihoods/HG002_GRCh38_1_22_v4.2.1_benchmark_noinconsistent_chr1.bed", 
    format = "BED") 

#change the seqnames to match the vcf file "chr1" to "NC_000001.11"
seqlevels(chr1_gr) <- gsub("chr1","NC_000001.11", seqlevels(chr1_gr))
```

### True homozygous reference CpG sites

Generate a truth set of homozygous reference CpG sites from reference CpG sites that do not overlap variants in the vcf file.

```{r}
#read the reference CpG sites and format them as GRanges
ref_CG <- import(
    "~/Documents/CpG_likelihoods/GCF_000001405.40_GRCh38.p14_genomic_CG_chr1.bed", 
    format = "BED"
) 
ref_CG <- ref_CG[strand(ref_CG) == "+"]
ref_CG <- subsetByOverlaps(ref_CG, chr1_gr, ignore.strand = T, type = "any")

# find CpG sites that do not overlap with any variants in the vcf file
vcf_gr<-GRanges(
    seqnames = vcf$ch, 
    ranges = IRanges(start = vcf$start-1, end = vcf$start+nchar(vcf$ref)+1))
ref_hom_CG<-subsetByOverlaps(
    ref_CG, vcf_gr, ignore.strand = T, invert = T, type = "any"
    )
```

### Remove neighbouring variants

We remove the variants where the neighboring position are also a variant as these neighboring variants can revert heterozygous CpG to homozygous CpG etc.

```{r}
vcf$next_pos <- c(vcf$start[-1], NA)
vcf$prev_pos <- c(NA, vcf$start[-nrow(vcf)])

# Identify rows with neighboring variants
neighbor_idx <- which(
  vcf$next_pos == vcf$start + 1 |
  vcf$prev_pos == vcf$start - 1
)

# Remove those rows
vcf_no_neighbours <- vcf[-neighbor_idx, ]
```

### True heterozygous reference CpG sites

We generate a truth set of heterozygous reference CpG sites by finding reference CpG sites that overlap heterozygous D > G or C > H variants in the vcf file.

```{r}
# Filter heterozygous variants (0/1 or 1/0),
# with C or G as reference and single-base ALT
vcf_het <- vcf_no_neighbours %>%
  filter(grepl("0/1|1/0", tag_info)) %>%
  filter(ref == "C" | ref == "G") %>%
  filter(nchar(alt) == 1)

# Adjust start for G reference alleles
vcf_het$start <- ifelse(
  vcf_het$ref == "G",
  vcf_het$start - 1,
  vcf_het$start
)

# Create GRanges object
vcf_het_gr <- GRanges(
  seqnames = vcf_het$ch,
  ranges = IRanges(
    start = vcf_het$start,
    end = vcf_het$start
  )
)

# Find overlaps with reference CpG sites
ref_het_CG <- subsetByOverlaps(
  ref_CG,
  vcf_het_gr,
  ignore.strand = TRUE,
  invert = FALSE,
  type = "any"
)
```

### True reference non-CpG sites

Generate a truth set of non-reference CpG sites by finding reference CpG sites that overlap homozygous D > G or C > H variants in the vcf file.

```{r}
# Filter homozygous variants (1/1),
# with C or G as reference and single-base ALT
vcf_non <- vcf_no_neighbours %>%
  filter(grepl("1/1", tag_info)) %>%
  filter(ref %in% c("C", "G")) %>%
  filter(nchar(alt) == 1)

# Adjust start position for G reference alleles
vcf_non$start <- ifelse(
  vcf_non$ref == "G",
  vcf_non$start - 1,
  vcf_non$start
)

# Create GRanges object
vcf_non_gr <- GRanges(
  seqnames = vcf_non$ch,
  ranges = IRanges(
    start = vcf_non$start,
    end = vcf_non$start
  )
)

# Find overlaps with reference CpG sites
ref_non_CG <- subsetByOverlaps(
  ref_CG,
  vcf_non_gr,
  ignore.strand = TRUE,
  invert = FALSE,
  type = "any"
)
```

### True homozygous non-reference CG loci

Generate a truth set of homozygous non-reference CpG sites by finding reference CH or DG sites that overlap homozygous D > G or C > H variants in the vcf file.

```{r}
# Read reference CH/DG sites and restrict to high-confidence regions
ref_CH <- import(
  "~/Documents/CpG_likelihoods/GCF_000001405.40_GRCh38.p14_genomic_CH_chr1.bed",
  format = "BED"
)

ref_CH <- subsetByOverlaps(
  ref_CH,
  chr1_gr,
  ignore.strand = TRUE,
  type = "any"
)

# Split into strands
ref_CH_plus  <- ref_CH[strand(ref_CH) == "+"]
ref_CH_minus <- ref_CH[strand(ref_CH) == "-"]

# PLUS STRAND: Homozygous H → G

vcf_hom_G <- vcf_no_neighbours %>%
  filter(grepl("1/1", tag_info)) %>%
  filter(ref != "G" & alt == "G") %>%
  filter(nchar(ref) == 1 & nchar(alt) == 1)

vcf_G_gr <- GRanges(
  seqnames = vcf_hom_G$ch,
  ranges = IRanges(
    start = vcf_hom_G$start - 1,
    end   = vcf_hom_G$start - 1
  )
)

vcf_CH_G <- subsetByOverlaps(
  ref_CH_plus,
  vcf_G_gr,
  ignore.strand = TRUE,
  invert = FALSE,
  type = "any"
)

# MINUS STRAND: Homozygous H → C

vcf_hom_C <- vcf_no_neighbours %>%
  filter(grepl("1/1", tag_info)) %>%
  filter(ref != "C" & alt == "C") %>%
  filter(nchar(ref) == 1 & nchar(alt) == 1)

vcf_C_gr <- GRanges(
  seqnames = vcf_hom_C$ch,
  ranges = IRanges(
    start = vcf_hom_C$start + 1,
    end   = vcf_hom_C$start + 1
  )
)

vcf_CH_C <- subsetByOverlaps(
  ref_CH_minus,
  vcf_C_gr,
  ignore.strand = TRUE,
  invert = FALSE,
  type = "any"
)

# Shift minus strand coordinates
vcf_CH_C <- shift(vcf_CH_C, -1)

# Combine results
ref_CH_hom <- c(vcf_CH_G, vcf_CH_C)
```

### True heterozygous non-reference CG loci

Generate a truth set of heterozygous non-reference CpG sites by finding reference CH or DG sites that overlap heterozygous D > G or C > H variants in the vcf file.

```{r}
# PLUS STRAND: Heterozygous H → G
vcf_het_G <- vcf_no_neighbours %>%
  filter(grepl("0/1|1/0", tag_info)) %>%
  filter(ref != "G" & alt == "G") %>%
  filter(nchar(ref) == 1, nchar(alt) == 1)

vcf_G_gr <- GRanges(
  seqnames = vcf_het_G$ch,
  ranges = IRanges(
    start = vcf_het_G$start - 1,
    end   = vcf_het_G$start - 1
  )
)

vcf_CH_G <- subsetByOverlaps(
  ref_CH_plus,
  vcf_G_gr,
  ignore.strand = TRUE,
  invert = FALSE,
  type = "any"
)

# MINUS STRAND: Heterozygous H → C
vcf_het_C <- vcf_no_neighbours %>%
  filter(grepl("0/1|1/0", tag_info)) %>%
  filter(ref != "C" & alt == "C") %>%
  filter(nchar(ref) == 1, nchar(alt) == 1)

vcf_C_gr <- GRanges(
  seqnames = vcf_het_C$ch,
  ranges = IRanges(
    start = vcf_het_C$start + 1,
    end   = vcf_het_C$start + 1
  )
)

vcf_CH_C <- subsetByOverlaps(
  ref_CH_minus,
  vcf_C_gr,
  ignore.strand = TRUE,
  invert = FALSE,
  type = "any"
)

# Shift minus strand coordinates
vcf_CH_C <- shift(vcf_CH_C, -1)

# Combine results
ref_CH_het <- c(vcf_CH_G, vcf_CH_C)
```

### True non-reference non-CpG loci

Generate a truth set of non-reference non-CpG sites by finding reference CH or DG sites that do not overlap with any variants in the vcf file.

```{r}
# Create GRanges object for all variants
vcf_gr <- GRanges(
  seqnames = vcf$ch,
  ranges = IRanges(
    start = vcf$start - 1,
    end   = vcf$start + nchar(vcf$ref) + 1
  )
)

# Find CH sites that do NOT overlap any variants
ref_CH_non <- subsetByOverlaps(
  ref_CH,
  vcf_gr,
  ignore.strand = TRUE,
  invert = TRUE,
  type = "any"
)
```

### Benmark data 
To benchmark the likelihood framework we have a truth set of 2.083.686 homozygous CpG loci, 26.352 heterozygous CpG loci and 14.805 non-CpG loci in the sample, which are all CpG loci in the reference genome. We also have a truth set of 16.891 homozygous CpG loci and 22.288 heterozygous CpG loci, and 84.991.174 non-CpG loci (CH/DG) in the sample, which are non-CpG loci in the reference genome.

```{r}
benchmark_data <- tibble(
  reference_CG = c("yes", "yes", "yes", "no", "no", "no"),
  type = c(
    "homozygous", "heterozygous", "non-CpG",
    "homozygous", "heterozygous", "non-CpG"
  ),
  n = c(
    length(ref_hom_CG),
    length(ref_het_CG),
    length(ref_non_CG),
    length(ref_CH_hom),
    length(ref_CH_het),
    length(ref_CH_non)
  )
)

benchmark_data
```

## Benchmarking
### Helper function to count overlaps between query and target GRanges objects

```{r}
# Helper: count overlaps

count_overlaps <- function(query_gr, targets) {
  c(
    length(subsetByOverlaps(query_gr, targets$ref_hom_CG, ignore.strand = TRUE)),
    length(subsetByOverlaps(query_gr, targets$ref_het_CG, ignore.strand = TRUE)),
    length(subsetByOverlaps(query_gr, targets$ref_non_CG, ignore.strand = TRUE)),
    length(subsetByOverlaps(query_gr, targets$ref_CH_hom, ignore.strand = TRUE)),
    length(subsetByOverlaps(query_gr, targets$ref_CH_het, ignore.strand = TRUE)),
    length(subsetByOverlaps(query_gr, targets$ref_CH_non, ignore.strand = TRUE))
  )
}
```

## Nanopore
### Read Nanopore data and build confusion matrix

```{r}
files <- list.files(
  "~/Documents/CpG_likelihoods/bedMethyls",
  pattern = "modkit",
  full.names = TRUE
)

mc <- read.bedMethyl(
  files[2],
  strandCollapse = TRUE,
  verbose = TRUE,
  output = "MethylCounts"
)

# Define truth sets (for cleaner calls)
targets <- list(
  ref_hom_CG = ref_hom_CG,
  ref_het_CG = ref_het_CG,
  ref_non_CG = ref_non_CG,
  ref_CH_hom = ref_CH_hom,
  ref_CH_het = ref_CH_het,
  ref_CH_non = ref_CH_non
)

# Build confusion matrix

# Total (truth)
tot <- sapply(targets, length)

# Different call types
raw <- count_overlaps(granges(mc), targets)

all <- count_overlaps(
  granges(mc[getCpGs(mc, type = "allCpG"), ]),
  targets
)

hom <- count_overlaps(
  granges(mc[getCpGs(mc, type = "homozygous"), ]),
  targets
)

het <- count_overlaps(
  granges(mc[getCpGs(mc, type = "heterozygous"), ]),
  targets
)

# Combine into matrix
conf_mat <- rbind(tot, raw, all, hom, het)

rownames(conf_mat) <- c(
  "benchmark loci",
  "unfiltered",
  "allCpG",
  "homozygous",
  "heterozygous"
)

colnames(conf_mat) <- c(
  "ref_homozygous",
  "ref_heterozygous",
  "ref_non-CpG",
  "homozygous",
  "heterozygous",
  "non-CpG"
)
```

### Plot Nanopore confussion matrix with unfiltered results included  

```{r}
# Prepare data for plotting

# Convert counts to frequencies
conf_mat_freq <- sweep(conf_mat[2:5, ], 2, conf_mat[1, ], "/")

# Convert to tidy format
conf_mat_freq_tbl <- as_tibble(conf_mat_freq, rownames = "Call") %>%
  pivot_longer(-Call, names_to = "Truth", values_to = "freq")

conf_mat_count_tbl <- as_tibble(conf_mat, rownames = "Call") %>%
  pivot_longer(-Call, names_to = "Truth", values_to = "n")

# Combine counts + frequencies
conf_mat_plot <- left_join(
  conf_mat_count_tbl,
  conf_mat_freq_tbl,
  by = c("Truth", "Call")
)

# Adjust frequencies for non-CpG classes (invert)
conf_mat_plot <- conf_mat_plot %>%
  mutate(
    freq = ifelse(
      Truth %in% c("ref_non-CpG", "non-CpG"),
      1 - freq,
      freq
    )
  )

# Set factor order
truth_levels <- c(
  "ref_homozygous",
  "ref_heterozygous",
  "ref_non-CpG",
  "homozygous",
  "heterozygous",
  "non-CpG"
)

call_levels <- c(
  "benchmark loci",
  "unfiltered",
  "allCpG",
  "homozygous",
  "heterozygous"
)

conf_mat_plot$Call  <- factor(conf_mat_plot$Call, levels = rev(call_levels))
conf_mat_plot$Truth <- factor(conf_mat_plot$Truth, levels = truth_levels)

# Plot heatmap
ggplot(conf_mat_plot, aes(x = Truth, y = Call, fill = freq)) +
  geom_tile() +
  geom_text(
    aes(
      label = n,
      color = ifelse(
        is.na(freq),
        "white",
        ifelse(freq < 0.5, "black", "white")
      )
    )
  ) +
  scale_color_identity() +
  scale_fill_gradientn(
    colors = RColorBrewer::brewer.pal(9, "Reds"),
    limits = c(0, 1)
  ) +
  scale_x_discrete(labels = function(x) gsub("^ref_", "", x)) +
  theme_minimal() +
  geom_vline(xintercept = 3.5, color = "white", size = 2) +
  geom_hline(yintercept = 3.5, color = "white", size = 1) +
  geom_hline(yintercept = 4.5, color = "white", size = 1) +
  labs(
    title = "",
    x = "True CpG status",
    y = "Called CpG status"
  ) +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12, angle = 55, hjust = 1),
    legend.position = "bottom",
    legend.key.height = unit(1.2, "cm"),
    legend.key.width  = unit(2.5, "cm"),
    legend.title = element_text(size = 14),
    legend.text  = element_text(size = 12)
  )
```

### Full table with sensitivity, and precision for Nanopore 

```{r}
# Sensitivity matrix
sens_mat <- sweep(conf_mat[2:5, ], 2, conf_mat[1, ], "/")

# Invert non-CpG classes
sens_mat[, c("ref_non-CpG", "non-CpG")] <-
  1 - sens_mat[, c("ref_non-CpG", "non-CpG")]

# Prepare matrix
conf_sub <- conf_mat[2:5, ]
ref_total    <- conf_sub[, 1] + conf_sub[, 2] + conf_sub[, 3]
nonref_total <- conf_sub[, 4] + conf_sub[, 5] + conf_sub[, 6]
row_total    <- rowSums(conf_sub)


# PPV calculations
ppv_ref <- c(
  (conf_sub["unfiltered", 1] + conf_sub["unfiltered", 2]) / ref_total["unfiltered"],
  (conf_sub["allCpG", 1]     + conf_sub["allCpG", 2])     / ref_total["allCpG"],
  conf_sub["homozygous", 1]  / ref_total["homozygous"],
  conf_sub["heterozygous", 2] / ref_total["heterozygous"]
)

ppv_nonref <- c(
  (conf_sub["unfiltered", 4] + conf_sub["unfiltered", 5]) / nonref_total["unfiltered"],
  (conf_sub["allCpG", 4]     + conf_sub["allCpG", 5])     / nonref_total["allCpG"],
  conf_sub["homozygous", 4]  / nonref_total["homozygous"],
  conf_sub["heterozygous", 5] / nonref_total["heterozygous"]
)

ppv_combined <- c(
  (conf_sub["unfiltered", 1] + conf_sub["unfiltered", 2] +
   conf_sub["unfiltered", 4] + conf_sub["unfiltered", 5]) / row_total["unfiltered"],

  (conf_sub["allCpG", 1] + conf_sub["allCpG", 2] +
   conf_sub["allCpG", 4] + conf_sub["allCpG", 5]) / row_total["allCpG"],

  (conf_sub["homozygous", 1] + conf_sub["homozygous", 4]) /
    row_total["homozygous"],

  (conf_sub["heterozygous", 2] + conf_sub["heterozygous", 5]) /
    row_total["heterozygous"]
)

# Final table
final_table <- as.data.frame(sens_mat) %>%
  rownames_to_column("Call") %>%
  mutate(
    ppv_ref = ppv_ref,
    ppv_nonref = ppv_nonref,
    ppv_combined = ppv_combined
  )

# Rename columns (short + consistent)
final_table <- final_table %>%
  rename(
    ref_hom = ref_homozygous,
    ref_het = ref_heterozygous,
    ref_nonCpG = `ref_non-CpG`,
    nonref_hom = homozygous,
    nonref_het = heterozygous,
    nonref_nonCpG = `non-CpG`
  )

# Format numbers
final_table_fmt <- final_table %>%
  mutate(across(-Call, ~ sprintf("%.5f", .)))

# Write output
write.table(
  final_table_fmt,
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)
```

## PacBio
### Read PacBio data and build confusion matrix

```{r}
files <- list.files(
  "~/Documents/CpG_likelihoods/bedMethyls",
  pattern = "modkit",
  full.names = TRUE
)

mc <- read.bedMethyl(
  files[3],
  strandCollapse = TRUE,
  verbose = TRUE,
  output = "MethylCounts"
)

# Define truth sets (for cleaner calls)
targets <- list(
  ref_hom_CG = ref_hom_CG,
  ref_het_CG = ref_het_CG,
  ref_non_CG = ref_non_CG,
  ref_CH_hom = ref_CH_hom,
  ref_CH_het = ref_CH_het,
  ref_CH_non = ref_CH_non
)

# Build confusion matrix

# Total (truth)
tot <- sapply(targets, length)

# Different call types
raw <- count_overlaps(granges(mc), targets)

all <- count_overlaps(
  granges(mc[getCpGs(mc, type = "allCpG"), ]),
  targets
)

hom <- count_overlaps(
  granges(mc[getCpGs(mc, type = "homozygous"), ]),
  targets
)

het <- count_overlaps(
  granges(mc[getCpGs(mc, type = "heterozygous"), ]),
  targets
)

# Combine into matrix
conf_mat <- rbind(tot, raw, all, hom, het)

rownames(conf_mat) <- c(
  "benchmark loci",
  "unfiltered",
  "allCpG",
  "homozygous",
  "heterozygous"
)

colnames(conf_mat) <- c(
  "ref_homozygous",
  "ref_heterozygous",
  "ref_non-CpG",
  "homozygous",
  "heterozygous",
  "non-CpG"
)
```

### Plot PacBio confussion matrix with unfiltered results included  

```{r}
# Prepare data for plotting

# Convert counts to frequencies
conf_mat_freq <- sweep(conf_mat[2:5, ], 2, conf_mat[1, ], "/")

# Convert to tidy format
conf_mat_freq_tbl <- as_tibble(conf_mat_freq, rownames = "Call") %>%
  pivot_longer(-Call, names_to = "Truth", values_to = "freq")

conf_mat_count_tbl <- as_tibble(conf_mat, rownames = "Call") %>%
  pivot_longer(-Call, names_to = "Truth", values_to = "n")

# Combine counts + frequencies
conf_mat_plot <- left_join(
  conf_mat_count_tbl,
  conf_mat_freq_tbl,
  by = c("Truth", "Call")
)

# Adjust frequencies for non-CpG classes (invert)
conf_mat_plot <- conf_mat_plot %>%
  mutate(
    freq = ifelse(
      Truth %in% c("ref_non-CpG", "non-CpG"),
      1 - freq,
      freq
    )
  )

# Set factor order
truth_levels <- c(
  "ref_homozygous",
  "ref_heterozygous",
  "ref_non-CpG",
  "homozygous",
  "heterozygous",
  "non-CpG"
)

call_levels <- c(
  "benchmark loci",
  "unfiltered",
  "allCpG",
  "homozygous",
  "heterozygous"
)

conf_mat_plot$Call  <- factor(conf_mat_plot$Call, levels = rev(call_levels))
conf_mat_plot$Truth <- factor(conf_mat_plot$Truth, levels = truth_levels)

# Plot heatmap
ggplot(conf_mat_plot, aes(x = Truth, y = Call, fill = freq)) +
  geom_tile() +
  geom_text(
    aes(
      label = n,
      color = ifelse(
        is.na(freq),
        "white",
        ifelse(freq < 0.5, "black", "white")
      )
    )
  ) +
  scale_color_identity() +
  scale_fill_gradientn(
    colors = RColorBrewer::brewer.pal(9, "Reds"),
    limits = c(0, 1)
  ) +
  scale_x_discrete(labels = function(x) gsub("^ref_", "", x)) +
  theme_minimal() +
  geom_vline(xintercept = 3.5, color = "white", size = 2) +
  geom_hline(yintercept = 3.5, color = "white", size = 1) +
  geom_hline(yintercept = 4.5, color = "white", size = 1) +
  labs(
    title = "",
    x = "True CpG status",
    y = "Called CpG status"
  ) +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12, angle = 55, hjust = 1),
    legend.position = "bottom",
    legend.key.height = unit(1.2, "cm"),
    legend.key.width  = unit(2.5, "cm"),
    legend.title = element_text(size = 14),
    legend.text  = element_text(size = 12)
  )
```

### Full table with sensitivity, and precision for PacBio

```{r}
# Sensitivity matrix
sens_mat <- sweep(conf_mat[2:5, ], 2, conf_mat[1, ], "/")

# Invert non-CpG classes
sens_mat[, c("ref_non-CpG", "non-CpG")] <-
  1 - sens_mat[, c("ref_non-CpG", "non-CpG")]

# Prepare matrix
conf_sub <- conf_mat[2:5, ]
ref_total    <- conf_sub[, 1] + conf_sub[, 2] + conf_sub[, 3]
nonref_total <- conf_sub[, 4] + conf_sub[, 5] + conf_sub[, 6]
row_total    <- rowSums(conf_sub)


# PPV calculations
ppv_ref <- c(
  (conf_sub["unfiltered", 1] + conf_sub["unfiltered", 2]) / ref_total["unfiltered"],
  (conf_sub["allCpG", 1]     + conf_sub["allCpG", 2])     / ref_total["allCpG"],
  conf_sub["homozygous", 1]  / ref_total["homozygous"],
  conf_sub["heterozygous", 2] / ref_total["heterozygous"]
)

ppv_nonref <- c(
  (conf_sub["unfiltered", 4] + conf_sub["unfiltered", 5]) / nonref_total["unfiltered"],
  (conf_sub["allCpG", 4]     + conf_sub["allCpG", 5])     / nonref_total["allCpG"],
  conf_sub["homozygous", 4]  / nonref_total["homozygous"],
  conf_sub["heterozygous", 5] / nonref_total["heterozygous"]
)

ppv_combined <- c(
  (conf_sub["unfiltered", 1] + conf_sub["unfiltered", 2] +
   conf_sub["unfiltered", 4] + conf_sub["unfiltered", 5]) / row_total["unfiltered"],

  (conf_sub["allCpG", 1] + conf_sub["allCpG", 2] +
   conf_sub["allCpG", 4] + conf_sub["allCpG", 5]) / row_total["allCpG"],

  (conf_sub["homozygous", 1] + conf_sub["homozygous", 4]) /
    row_total["homozygous"],

  (conf_sub["heterozygous", 2] + conf_sub["heterozygous", 5]) /
    row_total["heterozygous"]
)

# Final table
final_table <- as.data.frame(sens_mat) %>%
  rownames_to_column("Call") %>%
  mutate(
    ppv_ref = ppv_ref,
    ppv_nonref = ppv_nonref,
    ppv_combined = ppv_combined
  )

# Rename columns (short + consistent)
final_table <- final_table %>%
  rename(
    ref_hom = ref_homozygous,
    ref_het = ref_heterozygous,
    ref_nonCpG = `ref_non-CpG`,
    nonref_hom = homozygous,
    nonref_het = heterozygous,
    nonref_nonCpG = `non-CpG`
  )

# Format numbers
final_table_fmt <- final_table %>%
  mutate(across(-Call, ~ sprintf("%.5f", .)))

# Write output
write.table(
  final_table_fmt,
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)
```

## Modification levels for reference and non-reference CpGs 
### Define plot function

```{r}
make_mod_plot <- function(mc, ref_set, nonref_set) {
  
  # Subset data
  mc_ref <- subsetByOverlaps(mc, ref_set, ignore.strand = TRUE, type = "any")
  mc_nonref <- subsetByOverlaps(mc, nonref_set, ignore.strand = TRUE, type = "any")
  
  # Compute modification levels
  ref_mod <- (getMethylCounts(mc_ref, type="M") +
              getMethylCounts(mc_ref, type="H")) /
              getMethylCounts(mc_ref, type="Cov")
  
  nonref_mod <- (getMethylCounts(mc_nonref, type="M") +
                 getMethylCounts(mc_nonref, type="H")) /
                 getMethylCounts(mc_nonref, type="Cov")
  
  # KS test
  ks_res <- ks.test(ref_mod, nonref_mod)
  
  ks_label <- sprintf("KS = %.3f\np = %.1e",
                      ks_res$statistic,
                      ks_res$p.value)
  
  # Build dataframe
  mod_df <- data.frame(
    mod = c(ref_mod, nonref_mod),
    type = c(rep("reference", length(ref_mod)),
             rep("non-reference", length(nonref_mod)))
  )
  
  # Summary stats
  summary_stats <- mod_df %>%
    group_by(type) %>%
    summarise(mean_val = mean(mod, na.rm = TRUE), .groups = "drop")
  
  # Plot
  ggplot() +
    geom_half_violin(
      data = filter(mod_df, type == "reference"),
      aes(y = mod, fill = type),
      adjust = 3,
      side = "l",
      position = position_nudge(x = -0.005)
    ) +
    geom_half_violin(
      data = filter(mod_df, type == "non-reference"),
      aes(y = mod, fill = type),
      adjust = 3,
      side = "r",
      position = position_nudge(x = 0.005)
    ) +
    geom_segment(
      data = filter(summary_stats, type == "reference"),
      aes(x = -0.3, xend = -0.005, y = mean_val, yend = mean_val),
      size = 1
    ) +
    geom_segment(
      data = filter(summary_stats, type == "non-reference"),
      aes(x = 0.005, xend = 0.3, y = mean_val, yend = mean_val),
      size = 1
    ) +
    annotate(
      "text",
      x = 0.38,
      y = min(mod_df$mod, na.rm = TRUE) + 
          0.05 * (max(mod_df$mod, na.rm = TRUE) - min(mod_df$mod, na.rm = TRUE)),
      label = ks_label,
      hjust = 1,
      vjust = 0,
      size = 4
    ) +
    scale_fill_manual(values = c("reference" = "darkgrey",
                                 "non-reference" = "lightgrey")) +
    theme_bw() +
    theme(
      axis.title.y = element_blank()
    )
}
```

### Nanopore methylation plot
```{r}
files <- list.files(
  "~/Documents/CpG_likelihoods/bedMethyls",
  pattern = "modkit",
  full.names = TRUE
)

mc_nano <- read.bedMethyl(
  files[2],
  strandCollapse = TRUE,
  verbose = TRUE,
  output = "MethylCounts"
)

p1 <- make_mod_plot(
  mc_nano[getCpGs(mc_nano, type = "homozygous"), ],
  ref_hom_CG,
  ref_CH_hom
)

p2 <- make_mod_plot(
  mc_nano[getCpGs(mc_nano, type = "heterozygous"), ],
  ref_het_CG,
  ref_CH_het
)

p3 <- make_mod_plot(
  mc_nano[getCpGs(mc_nano, type = "allCpG"), ],
  c(ref_hom_CG, ref_het_CG),
  c(ref_CH_hom, ref_CH_het)
)
```

### PacBio methylation plot
```{r}
files <- list.files(
  "~/Documents/CpG_likelihoods/bedMethyls",
  pattern = "modkit",
  full.names = TRUE
)

mc_pac <- read.bedMethyl(
  files[2],
  strandCollapse = TRUE,
  verbose = TRUE,
  output = "MethylCounts"
)

p4 <- make_mod_plot(
  mc_pac[getCpGs(mc_pac, type = "homozygous"), ],
  ref_hom_CG,
  ref_CH_hom
)

p5 <- make_mod_plot(
  mc_pac[getCpGs(mc_pac, type = "heterozygous"), ],
  ref_het_CG,
  ref_CH_het
)

p6 <- make_mod_plot(
  mc_pac[getCpGs(mc_pac, type = "allCpG"), ],
  c(ref_hom_CG, ref_het_CG),
  c(ref_CH_hom, ref_CH_het)
)
```

### Final plot
```{r}
p1 <- p1 + labs(x = "")
p2 <- p2 + labs(x = "")
p3 <- p3 + labs(x = "")
p4 <- p4 + labs(x = "homozygous")
p5 <- p5 + labs(x = "heterozygous")
p6 <- p6 + labs(x = "allCpG")

combined_plot <- c(p1, p2, p3, p4, p5, p6) %>%
  wrap_plots(nrow = 2) +
  plot_layout(guides = "collect") &
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.text.x = element_text(size = 15),
    axis.text.y = element_text(size = 15),
    axis.title.x = element_text(size = 15),
    legend.position = "bottom",
    legend.key.size = unit(0.8, "cm"),
    legend.text = element_text(size = 15),
    legend.title = element_text(size = 15)
  )
#export pdf 10x10
```


# Silversides validation

### Read data 
```{r}
files <- list.files(
  "~/Desktop/BSseq_long-read/data/silversides_chr24/modkit",
  full.names = TRUE
)

mc_raw <- read.bedMethyl(
  files,
  strandCollapse = TRUE,
  verbose = TRUE,
  output = "MethylCounts"
)

G_all<-getCpGMatrix(mc_raw, allCpG = T)
Q_all<-getMaxLikelihoodMatrix(mc_raw, allCpG = T)

loci.idx <- which(
  DelayedMatrixStats::rowSums2(Q_all >= .99 & G_all==0) >= 2)
mc<-mc_raw[loci.idx]
```

### read reference CG loci and convert to GRanges 
```{r}
# Read reference CpG sites and filter chromosome + strand
ref_CG <- read_tsv(
  "~/Documents/Silversides/Mmenidia_refgenome_anchored.all_renamed_v2_CG.bed",
  col_names = c("seqnames", "start", "end", "X", "Y", "strand")
) %>%
  filter(seqnames == "Mme_chr24", strand == "+")

# Add zero-based positions (needed downstream)
ref_CG$start_zerobased <- ref_CG$start
ref_CG$end_zerobased   <- ref_CG$start

# Convert to GRanges
ref_CG_gr <- GRanges(
  seqnames = ref_CG$seqnames,
  ranges = IRanges(
    start = ref_CG$start,
    end   = ref_CG$end
  )
)
```

### Prepare functions and data for plotting
```{r}
# Count coverage ≥1 per column
count_coverage <- function(mc_obj) {
  sapply(1:9, function(i) {
    sum(getMethylCounts(mc_obj[, i], type = "Cov") >= 1)
  })
}

# Count overlaps for a given CpG type
count_overlaps_mc <- function(mc_obj, cpg_type, ref_gr, invert = FALSE) {
  sapply(1:9, function(i) {
    gr <- granges(mc_obj[, i][getCpGs(mc_obj[, i], type = cpg_type), ])
    length(
      subsetByOverlaps(
        gr,
        ref_gr,
        ignore.strand = TRUE,
        type = "any",
        invert = invert
      )
    )
  })
}
```

```{r}
# Coverage
unfiltered <- count_coverage(mc_raw)

# Reference CpG
all_ref <- count_overlaps_mc(mc, "allCpG", ref_CG_gr)
hom_ref <- count_overlaps_mc(mc, "homozygous", ref_CG_gr)
het_ref <- count_overlaps_mc(mc, "heterozygous", ref_CG_gr)

# Non-reference CpG
all_nonref <- count_overlaps_mc(mc, "allCpG", ref_CG_gr, invert = TRUE)
hom_nonref <- count_overlaps_mc(mc, "homozygous", ref_CG_gr, invert = TRUE)
het_nonref <- count_overlaps_mc(mc, "heterozygous", ref_CG_gr, invert = TRUE)
```

```{r}
mat <- matrix(
  c(
    unfiltered,
    hom_ref, het_ref, all_ref,
    hom_nonref, het_nonref, all_nonref
  ),
  ncol = 7,
  byrow = FALSE
)

rownames(mat) <- c("AA1","AA2","AA3","RA1","RA2","RA3","RR1","RR2","RR3")

colnames(mat) <- c(
  "unfiltered",
  "ref_homozygous",
  "ref_heterozygous",
  "ref_allCpG",
  "homozygous",
  "heterozygous",
  "allCpG"
)

mat
```

### Plot Silversides data with counts and frequencies
```{r}
# Convert matrix to long format
mat_tbl <- as_tibble(mat, rownames = "sample") %>%
  pivot_longer(-sample, names_to = "Call", values_to = "n")

# Compute frequencies
mat_freq <- cbind(
  ref_homozygous   = mat[, 1] / (mat[, 1] + mat[, 4]),
  ref_heterozygous = mat[, 2] / (mat[, 2] + mat[, 5]),
  ref_allCpG       = mat[, 3] / (mat[, 3] + mat[, 6]),
  homozygous       = mat[, 4] / (mat[, 1] + mat[, 4]),
  heterozygous     = mat[, 5] / (mat[, 2] + mat[, 5]),
  allCpG           = mat[, 6] / (mat[, 3] + mat[, 6])
)

rownames(mat_freq) <- rownames(mat)
colnames(mat_freq) <- c(
  "ref_homozygous",
  "ref_heterozygous",
  "ref_allCpG",
  "homozygous",
  "heterozygous",
  "allCpG"
)

mat_freq_tbl <- as_tibble(mat_freq, rownames = "sample") %>%
  pivot_longer(-sample, names_to = "Call", values_to = "freq")

# Combine counts + frequencies
mat_plot <- left_join(mat_tbl, mat_freq_tbl, by = c("sample", "Call"))

# Remove frequency for unfiltered
mat_plot$freq[mat_plot$Call == "unfiltered"] <- NA

# Factor ordering
call_levels <- c(
  "unfiltered",
  "ref_homozygous", "ref_heterozygous", "ref_allCpG",
  "homozygous", "heterozygous", "allCpG"
)

sample_levels <- rev(c(
  "RR1", "RR2", "RR3",
  "RA1", "RA2", "RA3",
  "AA1", "AA2", "AA3"
))

mat_plot$Call   <- factor(mat_plot$Call, levels = call_levels)
mat_plot$sample <- factor(mat_plot$sample, levels = sample_levels)

# Plot
ggplot(mat_plot, aes(x = Call, y = sample, fill = freq)) +
  geom_tile() +
  geom_text(
    aes(
      label = n,
      color = ifelse(is.na(freq) | freq < 0.5, "black", "white")
    )
  ) +
  scale_color_identity() +
  scale_fill_gradientn(
    colors = RColorBrewer::brewer.pal(9, "Blues"),
    limits = c(0, 1),
    na.value = "grey"
  ) +
  scale_x_discrete(labels = function(x) gsub("^ref_", "", x)) +
  theme_minimal() +
  geom_vline(xintercept = c(1.5, 4.5), color = "white", size = 2) +
  geom_vline(xintercept = c(3.5, 6.5), color = "white", size = 1) +
  geom_hline(yintercept = c(3.5, 6.5), color = "white", size = 2) +
  labs(
    title = "",
    x = "",
    y = "",
    fill = "Reference to non-reference ratio"
  ) +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.position = "bottom",
    legend.key.height = unit(1.2, "cm"),
    legend.key.width  = unit(2.5, "cm"),
    legend.title = element_text(size = 14),
    legend.text  = element_text(size = 12)
  )
```

### Prepare silversides methylation plot
```{r}
make_mod_plot <- function(mc, ref_set) {
  
  # Subset data
  mc_ref <- subsetByOverlaps(mc, ref_set, ignore.strand = TRUE, type = "any")
  mc_nonref <- subsetByOverlaps(mc, ref_set, ignore.strand = TRUE, type = "any", invert = TRUE)
  
  # Compute modification levels
  ref_mod <- (getMethylCounts(mc_ref, type="M") +
              getMethylCounts(mc_ref, type="H")) /
              getMethylCounts(mc_ref, type="Cov")
  
  nonref_mod <- (getMethylCounts(mc_nonref, type="M") +
                 getMethylCounts(mc_nonref, type="H")) /
                 getMethylCounts(mc_nonref, type="Cov")
  
  # KS test
  ks_res <- ks.test(ref_mod, nonref_mod)
  
  ks_label <- sprintf("KS = %.3f\np = %.1e",
                      ks_res$statistic,
                      ks_res$p.value)
  
  # Build dataframe
  mod_df <- data.frame(
    mod = c(ref_mod, nonref_mod),
    type = c(rep("reference", length(ref_mod)),
             rep("non-reference", length(nonref_mod)))
  )
  
  # Summary stats
  summary_stats <- mod_df %>%
    group_by(type) %>%
    summarise(mean_val = mean(mod, na.rm = TRUE), .groups = "drop")
  
  # Plot
  ggplot() +
    geom_half_violin(
      data = filter(mod_df, type == "reference"),
      aes(y = mod, fill = type),
      adjust = 3,
      side = "l",
      position = position_nudge(x = -0.005)
    ) +
    geom_half_violin(
      data = filter(mod_df, type == "non-reference"),
      aes(y = mod, fill = type),
      adjust = 3,
      side = "r",
      position = position_nudge(x = 0.005)
    ) +
    geom_segment(
      data = filter(summary_stats, type == "reference"),
      aes(x = -0.3, xend = -0.005, y = mean_val, yend = mean_val),
      size = 1
    ) +
    geom_segment(
      data = filter(summary_stats, type == "non-reference"),
      aes(x = 0.005, xend = 0.3, y = mean_val, yend = mean_val),
      size = 1
    ) +
    annotate(
      "text",
      x = 0.38,
      y = min(mod_df$mod, na.rm = TRUE) + 
          0.05 * (max(mod_df$mod, na.rm = TRUE) - min(mod_df$mod, na.rm = TRUE)),
      label = ks_label,
      hjust = 1,
      vjust = 0,
      size = 4
    ) +
    scale_fill_manual(values = c("reference" = "darkgrey",
                                 "non-reference" = "lightgrey")) +
    theme_bw() +
    theme(
      axis.title.y = element_blank()
    )
}
```

### Silversides methylation plot
```{r}
samples <-c("AA1","AA2","AA3","RA1","RA2", "RA3", "RR1", "RR2", "RR3")

p1 <- make_mod_plot(
  mc[,1][getCpGs(mc[,1], type = "allCpG"), ],
  ref_set = ref_CG_gr
) + labs(x = samples[1])

p2<- make_mod_plot(
  mc[,2][getCpGs(mc[,2], type = "allCpG"), ],
  ref_set = ref_CG_gr
) + labs(x = samples[2])

p3<- make_mod_plot(
  mc[,3][getCpGs(mc[,3], type = "allCpG"), ],
  ref_set = ref_CG_gr
) + labs(x = samples[3])

p4<- make_mod_plot(
  mc[,4][getCpGs(mc[,4], type = "allCpG"), ],
  ref_set = ref_CG_gr
) + labs(x = samples[4])

p5<- make_mod_plot(
  mc[,5][getCpGs(mc[,5], type = "allCpG"), ],
  ref_set = ref_CG_gr
) + labs(x = samples[5])

p6<- make_mod_plot(
  mc[,6][getCpGs(mc[,6], type = "allCpG"), ],
  ref_set = ref_CG_gr
) + labs(x = samples[6])

p7<- make_mod_plot(
  mc[,7][getCpGs(mc[,7], type = "allCpG"), ],
  ref_set = ref_CG_gr
) + labs(x = samples[7])

p8<- make_mod_plot(
  mc[,8][getCpGs(mc[,8], type = "allCpG"), ],
  ref_set = ref_CG_gr
) + labs(x = samples[8])

p9<- make_mod_plot(
  mc[,9][getCpGs(mc[,9], type = "allCpG"), ],
  ref_set = ref_CG_gr
) + labs(x = samples[9])
```

```{r}
combined_plot_silver <-c(p7, p8, p9, p4, p5, p6, p1, p2, p3) %>%
  wrap_plots(nrow = 3) +
  plot_layout(guides = "collect") &
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.text.x = element_text(size = 15),
    axis.text.y = element_text(size = 15),
    axis.title.x = element_text(size = 15),
    legend.position = "bottom",
    legend.key.size = unit(0.8, "cm"),
    legend.text = element_text(size = 15),
    legend.title = element_text(size = 15)
  )
combined_plot_silver
```

